---
title: "Citibike Bike Usage"
author: "Justin Applefield"
date: "12/03/2021"
output: 
  html_document:
    code_folding: hide  
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

```{r, cache=FALSE}
# Loading the Libraries
library(tidyverse)
library(ggplot2)
library(gganimate)
library(leaflet)
```

#Citibike Team # Group Project  
```{r}
# Observing the Data
citi <- read_csv("citicleaned.csv")

library(janitor)
citi <- clean_names(citi)

citi$start_station_name <- as.factor(citi$start_station_name)
citi$start_station_id <- as.factor(citi$start_station_id)
citi$end_station_name <- as.factor(citi$end_station_name)
citi$end_station_id <- as.factor(citi$end_station_id)
citi$usertype <- as.factor(citi$usertype)
citi$gender <- as.factor(citi$gender)
citi$dayid <- as.factor(citi$dayid)

# citi$month <- as.factor(citi$month)

str(citi)
```

## Bike Usage Analysis 

I'm seeking to answer the following questions:

### How does bike usage change over time?

#### Number of Rides per Day per bike
```{r}
daily_usage <- citi %>%
  group_by(dayid, day, bikeid) %>%
  summarize(num_rides = n())
```

#### Average number of 

### How does bike usage change month to month?

#### Number of Rides per Month
```{r}
monthly_usage <- citi %>%
  group_by(month) %>%
  summarize(num_rides = n())
```

### How does bike usage change between weekdays and weekends?
```{r}
weekday_vs_weekend <- daily_usage %>%
  group_by(dayid) %>%
  summarize(avg_rides = mean(num_rides))

weekday_vs_weekend_by_month <- citi %>%
  group_by(month, dayid) %>%
  summarize(avg_rides = mean(n()))

gif <- weekday_vs_weekend_by_month %>% ggplot(aes(x = month, y = avg_rides, color = dayid)) + 
  scale_x_continuous(breaks = seq(1, 12, by = 1)) +
  geom_line() + transition_reveal(month)
anim_save("g.gif", gif)
```

### How does bike usage change season to season?
```{r}

```

### What bikes are used the most?

#### Number of Rides per Bike
```{r}
bike_usage <- citi %>%
  group_by(bikeid) %>%
  summarize(num_rides = n())
```

### Do certain bikes make it back to the same stations?
```{r}
bike_return <- citi %>%
  group_by(end_station_id, bikeid) %>%
  summarize(num_returns = n()) %>%
  filter(num_returns > 2)
```

### How many trips do certain bikes take per day?
```{r}
bike_daily_usage <- citi %>%
  group_by(day, bikeid) %>%
  summarize(num_rides = n())
```

### Where do bikes travel (on a map)?
```{r}

```


