---
title: "Citibike Bike Usage"
author: "Justin Applefield"
date: "12/03/2021"
output: 
  html_document:
    code_folding: hide  
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

```{r, cache=FALSE}
# Loading the Libraries
library(tidyverse)
library(ggplot2)
library(gganimate)
library(leaflet)
library(janitor)
```

#Citibike Team # Group Project  
```{r}
# Observing the Data
citi <- read_csv("citicleaned.csv")
# citi <- read_csv("citiall.csv")

citi <- clean_names(citi)

citi$start_station_name <- as.factor(citi$start_station_name)
citi$start_station_id <- as.factor(citi$start_station_id)
citi$end_station_name <- as.factor(citi$end_station_name)
citi$end_station_id <- as.factor(citi$end_station_id)
citi$usertype <- as.factor(citi$usertype)
citi$gender <- as.factor(citi$gender)
citi$dayid <- as.factor(citi$dayid)


str(citi)
```

## Bike Usage Analysis 

I'm seeking to answer the following questions:

### How does bike usage change over time?

#### Number of Rides per Day per bike
```{r}
daily_usage <- citi %>%
  group_by(dayid, day, bikeid) %>%
  summarize(num_rides = n())
```

```{r}
daily_usage %>% 
  ggplot(aes(x = num_rides)) + geom_histogram()

daily_usage %>%
  filter(num_rides > 2) %>%
  nrow()

daily_usage %>%
  filter(num_rides > 3) %>%
  nrow()
```


#### Average number of rides per given bike
```{r}
avg_bike_usage <- daily_usage %>%
  group_by(bikeid) %>%
  summarize(avg_num_rides = mean(num_rides)) %>%
  arrange(desc(avg_num_rides))
```

```{r}
avg_bike_usage %>%
  ggplot(aes(x=avg_num_rides)) + geom_histogram()
```


### How does bike usage change month to month?

#### Number of Rides per bike per Month
```{r}
monthly_usage <- citi %>%
  group_by(month, bikeid) %>%
  summarize(num_rides = n())
```

```{r}
monthly_usage %>%
  ggplot(aes(x=num_rides)) + geom_histogram() + facet_wrap(~month)
```


### How does bike usage change between weekdays and weekends?

#### At all?
```{r}
weekday_vs_weekend <- daily_usage %>%
  group_by(dayid, bikeid) %>%
  summarize(avg_rides = mean(num_rides)) %>%
  ungroup()

# weekday_usage <- weekday_vs_weekend %>%
#   filter(dayid == "Weekday") %>%
#   select(bikeid, avg_rides) %>%
#   rename(avg_weekday_rides = avg_rides)
# 
# weekend_usage <- weekday_vs_weekend %>%
#   filter(dayid == "Weekend") %>%
#   select(bikeid, avg_rides) %>%
#   rename(avg_weekend_rides = avg_rides)
# 
# weekday_vs_weekend <- weekday_usage %>%
#   full_join(weekend_usage) %>%
#   mutate_all(~ replace(., is.na(.), 0))
```

```{r}

```

#### Month to month?
```{r}
daily_usage_by_month <- citi %>%
  group_by(dayid, month, day, bikeid) %>%
  summarize(num_rides = n())

weekday_vs_weekend_by_month <- daily_usage_by_month %>%
  group_by(dayid, month, bikeid) %>%
  summarize(avg_rides = mean(num_rides)) %>%
  ungroup()

# weekday_usage_by_month <- weekday_vs_weekend_by_month %>%
#   filter(dayid == "Weekday") %>%
#   select(bikeid, avg_rides, month) %>%
#   rename(avg_weekday_rides = avg_rides)
# 
# weekend_usage_by_month <- weekday_vs_weekend_by_month %>%
#   filter(dayid == "Weekend") %>%
#   select(bikeid, avg_rides, month) %>%
#   rename(avg_weekend_rides = avg_rides)
# 
# weekday_vs_weekend_by_month <- weekday_usage_by_month %>%
#   full_join(weekend_usage_by_month) %>%
#   mutate_all(~ replace(., is.na(.), 0))
```

### What bikes are used the most?

#### Number of Rides per Bike
```{r}
bike_usage <- citi %>%
  group_by(bikeid) %>%
  summarize(num_rides = n()) %>%
  arrange(desc(num_rides))
```

```{r}
bike_usage %>%
  ggplot(aes(x=num_rides)) + geom_histogram()
```


### Do certain bikes make it back to the same stations?

```{r}
station_info <- citi %>% group_by(end_station_id) %>% arrange(end_station_id) %>% filter(row_number() == 1) %>% ungroup() %>% distinct(end_station_id, end_station_name, end_station_latitude, end_station_longitude)
```

#### Ever?
```{r}
bike_return_ever <- citi %>%
  group_by(end_station_id, bikeid) %>%
  summarize(num_returns = n()) %>%
  arrange(desc(num_returns)) %>%
  left_join(station_info)
```

```{r}

```


#### On the same day?
```{r}
bike_return_day <- citi %>%
  group_by(day, end_station_id, bikeid) %>%
  summarize(num_returns = n()) %>%
  arrange(num_returns) %>%
  left_join(station_info)
```

```{r}
  leaflet(bike_return_day) %>% 
  setView(-74, 40.75, zoom = 11.5) %>%
  addMarkers(lng = ~end_station_longitude, lat = ~end_station_latitude, 
                   popup = paste(~bikeid, ~end_station_name, "<br>", ~num_returns))
```


### How many trips do certain bikes take per day?
```{r}
bike_daily_usage <- citi %>%
  group_by(day, bikeid) %>%
  summarize(num_rides = n())
```

```{r}

```


### Where do bikes travel (on a map)?
```{r}

```


