---
title: "Citibike Bike Usage"
author: "Justin Applefield"
date: "12/03/2021"
output: 
  html_document:
    code_folding: hide  
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

```{r, cache=FALSE}
# Loading the Libraries
library(tidyverse)
library(ggplot2)
library(gganimate)
library(leaflet)
library(janitor)
```

#Citibike Team # Group Project  
```{r}
# Observing the Data
citi <- read_csv("citicleaned.csv")
# citi <- read_csv("citiall.csv")

citi <- clean_names(citi)

citi$start_station_name <- as.factor(citi$start_station_name)
citi$start_station_id <- as.factor(citi$start_station_id)
citi$end_station_name <- as.factor(citi$end_station_name)
citi$end_station_id <- as.factor(citi$end_station_id)
citi$usertype <- as.factor(citi$usertype)
citi$gender <- as.factor(citi$gender)
citi$dayid <- as.factor(citi$dayid)


str(citi)
```

## Bike Usage Analysis 

I'm seeking to answer the following questions:

### How does bike usage change over time?

#### Number of Rides per Day per bike
```{r}
daily_usage <- citi %>%
  group_by(dayid, day, bikeid) %>%
  summarize(num_rides = n())
```

```{r}
daily_usage %>%
  ggplot(aes(x = num_rides)) +
  geom_histogram()

daily_usage %>%
  filter(num_rides > 2) %>%
  nrow()

daily_usage %>%
  filter(num_rides > 3) %>%
  nrow()
```


#### Average number of rides per given bike
```{r}
avg_bike_usage <- daily_usage %>%
  group_by(bikeid) %>%
  summarize(avg_num_rides = mean(num_rides)) %>%
  arrange(desc(avg_num_rides))
```

```{r}
avg_bike_usage %>%
  ggplot(aes(x = avg_num_rides)) +
  geom_histogram()
```


### How does bike usage change month to month?

#### Number of Rides per bike per Month
```{r}
monthly_usage <- citi %>%
  group_by(month, bikeid) %>%
  summarize(num_rides = n())
```

```{r}
monthly_usage %>%
  ggplot(aes(x = num_rides)) +
  geom_histogram() +
  facet_wrap(~month)
```


### How does bike usage change between weekdays and weekends?

#### At all?
```{r}
weekday_vs_weekend <- daily_usage %>%
  group_by(dayid, bikeid) %>%
  summarize(avg_rides = mean(num_rides)) %>%
  ungroup()

# weekday_usage <- weekday_vs_weekend %>%
#   filter(dayid == "Weekday") %>%
#   select(bikeid, avg_rides) %>%
#   rename(avg_weekday_rides = avg_rides)
#
# weekend_usage <- weekday_vs_weekend %>%
#   filter(dayid == "Weekend") %>%
#   select(bikeid, avg_rides) %>%
#   rename(avg_weekend_rides = avg_rides)
#
# weekday_vs_weekend <- weekday_usage %>%
#   full_join(weekend_usage) %>%
#   mutate_all(~ replace(., is.na(.), 0))
```

```{r}

```

#### Month to month?
```{r}
daily_usage_by_month <- citi %>%
  group_by(dayid, month, day, bikeid) %>%
  summarize(num_rides = n())

weekday_vs_weekend_by_month <- daily_usage_by_month %>%
  group_by(dayid, month, bikeid) %>%
  summarize(avg_rides = mean(num_rides)) %>%
  ungroup()

# weekday_usage_by_month <- weekday_vs_weekend_by_month %>%
#   filter(dayid == "Weekday") %>%
#   select(bikeid, avg_rides, month) %>%
#   rename(avg_weekday_rides = avg_rides)
#
# weekend_usage_by_month <- weekday_vs_weekend_by_month %>%
#   filter(dayid == "Weekend") %>%
#   select(bikeid, avg_rides, month) %>%
#   rename(avg_weekend_rides = avg_rides)
#
# weekday_vs_weekend_by_month <- weekday_usage_by_month %>%
#   full_join(weekend_usage_by_month) %>%
#   mutate_all(~ replace(., is.na(.), 0))
```

### What bikes are used the most?

#### Number of Rides per Bike
```{r}
bike_usage <- citi %>%
  group_by(bikeid) %>%
  summarize(num_rides = n()) %>%
  arrange(desc(num_rides))
```

```{r}
bike_usage %>%
  ggplot(aes(x = num_rides)) +
  geom_histogram()
```


### Do certain bikes make it back to the same stations?

```{r}
station_info <- citi %>%
  group_by(end_station_id) %>%
  arrange(end_station_id) %>%
  filter(row_number() == 1) %>%
  ungroup() %>%
  distinct(end_station_id, end_station_name, latitude = end_station_latitude, longitude = end_station_longitude)
```

#### Ever?
```{r}
bike_return_ever <- citi %>%
  group_by(end_station_id, bikeid) %>%
  summarize(num_returns = n()) %>%
  arrange(desc(num_returns)) %>%
  left_join(station_info) %>%
  head(n = 20)
```

```{r}
leaflet(bike_return_ever) %>%
  addTiles() %>%
  setView(-74, 40.75, zoom = 11.5) %>%
  addMarkers(
    lng = ~longitude, lat = ~latitude,
    popup = paste(bike_return_ever$bikeid, bike_return_ever$end_station_name, "<br>", bike_return_ever$num_returns)
  )
```


#### On the same day?
```{r}
bike_return_day <- citi %>%
  group_by(day, end_station_id, bikeid) %>%
  summarize(num_returns = n()) %>%
  arrange(desc(num_returns)) %>%
  left_join(station_info) %>%
  head(n = 20)
```

```{r}
leaflet(bike_return_day) %>%
  addTiles() %>%
  setView(-74, 40.75, zoom = 11.5) %>%
  addMarkers(
    lng = ~longitude, lat = ~latitude,
    popup = paste(bike_return_day$bikeid, bike_return_day$end_station_name, "<br>", bike_return_day$num_returns)
  )
```


### How do the most used bikes move on their active days?
```{r}
bike_daily_usage <- citi %>%
  arrange(starttime) %>%
  group_by(day, bikeid) %>%
  mutate(num_rides = n(), first_ride = row_number() == 1, last_ride = row_number() == n()) %>%
  arrange(desc(num_rides)) %>%
  filter(first_ride | last_ride) %>%
  head(n = 20)

active_firsts <- bike_daily_usage %>%
  filter(first_ride) %>%
  select(bikeid, day, start_station_id, start_station_latitude, start_station_longitude, start_station_name)

active_lasts <- bike_daily_usage %>%
  filter(last_ride) %>%
  select(bikeid, day, end_station_id, end_station_latitude, end_station_longitude, end_station_name)

bike_daily_usage <- active_firsts %>%
  full_join(active_lasts)
```

```{r}
library(leaflet.minicharts)
leaflet(bike_daily_usage) %>%
  addTiles() %>%
  setView(-74, 40.75, zoom = 11.5) %>%
  addFlows(
    lng0 = bike_daily_usage$start_station_longitude,
    lng1 = bike_daily_usage$end_station_longitude,
    lat0 = bike_daily_usage$start_station_latitude,
    lat1 = bike_daily_usage$end_station_latitude,
    maxThickness = 3
  )
```


